Introduction to Structural Causal Models

### 2021-05-21 presentation_video.Rmd
video

---
title: "Introduction to Structural Causal Models"
author: "Daniel Saggau"
date: "5/11/2021"
citation_package: --biblatex
bibliography: [ref.bib]
output: 
  beamer_presentation:
    theme: Madrid
header-includes:
- \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
- \usecolortheme[named=UBCblue]{structure}
---

@@

* Foundations of SCMs 
  * Assumptions 
  * Comparing Causal Tools 
* Pearl's Causal Hierachy 
  * Prediction 
  * Intervention 
  * Counterfactuals 
  * Implications 
* Graphical Models 
  * Features 
  * Implications
* Causality and Time

# Foundations of SCMs (1)

- System of equations with functions
- Assignment equation ':=' rather than regular equation '=' 
- Nonparametric SEM 
- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations (symmetry)
- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.

**Error:* **
Regression: Omittable outside factor 
SCM/SEM; Latent influential factor that is pivotal for the model specification but not observable

@@

# Assumptions in Causality

@@

- noise terms independent
- mechanisms independent (other variables invariant)(local changes)
- change in distribution stems from change in mechanism

- Causal Markov Condition allows to factorize 
- Each Conditional is a causal mechanism

Source: [@peters_elements_2017]

@@

**Established Conditions:**

Definition: **SUTVA**

'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]

Definition: **Consistency**

The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]

Definition: **Ignorability**

The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
(perfect RCT)

- First tow hold for SCM counterfactuals 
- third not testable but can check via backdoor criterion in SCM

@@

@@

- ML, BN and regression are at the lowest level in the causal hierarchy
- Prediction methods demand the least information and depend on association alone
- Associational methods ignore external changes outside of our data.
- intervention distribution has information on these external changes.
- intervention distribution is only defined in high level causal methods.

@@

- The second query deals with interventions
- Mathematical Tool: do-calculus
- The do-calculus enables us to study the manipulation of parent nodes
- There are various types of intervention
- One example is **atomic intervention,** where we set a variable to a constant
- In **policy intervention** we specify a different function for an equation
- Off-policy intervention models different intervention that is not in our historical data
- CBN , MDP and reinforcement learning model intervention.

@@

- by replacing function with a constant

$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=600$$
$$C_2:= f_{c_2}(a,N_2)$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

- Intervention by replacing function with a different conditional probability

$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=f(\pi)$$
$$C_2:= f_{c_2}(a,N_2)$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

@@

- Nodes -> Variables (endogenous/exogenous)
- Edges -> relationship (equations)
- Parents/Ancestors/Descendents
- No need to specify exact parametric shape
- highlight colliders 
- estimation back door criterion 
- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.)

@@

@@

$$
\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
$$

@@

@@

- Mediation Analysis 
- PO-Framework
- Causal Algorithms
- Causal Constraints Model

# Conclusions

- Assosicational learning is easy to model because of lower information neccessary
- but not always appropiate in high stake settings 
- SCM as expressive simulator for causal modelling outperforms many other models
- computational advantage casting causal model as system of assignment equations 
- Enables modelling of higher order concepts like counterfactuals 
- Extensions through differential equations for concise modelling of time

\newpage 
# References

<div id="refs"></div>
\newpage

>>>>

---
title: "Introduction to Structural Causal Models"
author: "Daniel Saggau"
date: "5/11/2021"
citation_package: --biblatex
bibliography: [ref.bib]
csl: ieee.csl
output: 
  beamer_presentation:
    theme: Madrid
header-includes:
- \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
- \usecolortheme[named=UBCblue]{structure}
---

@@

* Foundations of SCMs 
  * Assumptions 
  * Comparing Causal Tools 
* Pearl's Causal Hierachy 
  * Prediction 
  * Intervention 
  * Counterfactuals 
* Graphical Models 
* Causality and Time

# Foundations of SCMs

- Nonparametric SEM 
- System of equations with functions
- Assignments ':=' (non-symmetric) instead of algebraic equation '=' (symmetry)
- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.
**Error in Regression:** Omittable outside factor 
**Error in SCM/SEM:** Latent (influential) factor that is pivotal for the model

@@

# Assumptions in Causality (1)

@@

- Noise terms independent $(N_c,N_e)$
- Mechanisms independent (other variables invariant)(local changes)
- Change in distribution stems from change in mechanism 
- Causal Markov Condition

@@

 **SUTVA**; 'The treatment that one unit receives does not change the effect of treatment for any other unit.'

**Consistency**: The outcome Y agrees with the potential outcome corresponding to the treatment indicator.'

**Ignorability**: The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. (perfect RCT)

- First tow hold for SCM counterfactuals 
- third not testable but can check via backdoor criterion in SCM [@hardtrecht]

@@

* Foundations of SCMs 
  * Assumptions 
  * Comparing Causal Tools 
* **Pearl's Causal Hierachy**
  * Prediction 
  * Intervention 
  * Counterfactuals 
* Graphical Models 
* Causality and Time

@@

- ML, BN and regression are at the lowest level in the causal hierarchy
- Prediction methods demand the least information and depend on association alone
- Association-based methods ignore external changes outside of our data
- Intervention distribution needed for higher level information

@@

- Mathematical Tool: do-calculus
- The do-calculus enables us to study the manipulation of parent nodes
- **Atomic intervention:** where we set a variable to a constant
- **Policy intervention:** we specify a different function for an equation
- CBN , MDP and reinforcement learning model intervention.

@@

- replacing function with a constant

$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=600$$
$$C_2:= f_{c_2}(a,N_{c_2})$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

- replacing function with a different conditional probability

$$C_1:= f_{c_1}(p,N_{c_1}) \rightarrow C_1:=f(\pi)$$
$$C_2:= f_{c_2}(a,N_{c_2})$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

* Foundations of SCMs 
  * Assumptions 
  * Comparing Causal Tools 
* Pearl's Causal Hierachy
  * Prediction 
  * Intervention 
  * Counterfactuals 
* **Graphical Models**
* Causality and Time

@@

- Nodes -> Variables (endogenous/exogenous)
- Edges -> relationship (equations)
- Parents/Ancestors/Descendents
- No need to specify exact parametric shape
- Highlight colliders 
- Estimation back door criterion 
- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.)

@@

* Foundations of SCMs 
  * Assumptions 
  * Comparing Causal Tools 
* Pearl's Causal Hierachy
  * Prediction 
  * Intervention 
  * Counterfactuals 
* Graphical Models
* **Causality and Time**

@@

Derivative of function x with respect to time t:

$$
\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
$$
Value of Function at time t + dt:

@@

- dependence on prior time point and change in time contribute to the value at time point

@@

- Mediation Analysis 
- PO-Framework
- Causal Algorithms
- IV-Estimation
- Causal Constraints Model 
- Causal Trees

# Concluding Remarks

- Assosicational learning is easy to model because of lower information neccessary
- but not always appropiate in high stake settings 
- SCM as simulator for causal modelling, entailing a lot of information (DGP, intervention distribution)
- Note: *"Garbage in, Garbage out"*
- computational advantage casting causal model as system of assignments 
- Enables modelling of higher order concepts like counterfactuals 
- Extensions through differential equations for concise modelling of time

\newpage 
# References
\tiny

<div id="refs"></div>
\normalsize
\newpage
